---
title: "aoristic compare"
output: html_document
---

### Comparison of execution times of `archSeries::aorist` and my version of the same analysis; `my_aorist`.
I performed this purely out of interest in how the packages approach (`data.table` and `for` loops) and my approach (`data.table::foverlaps()` and `dplyr`).  The key difference to these approaches is that `archSeries::aorist()` uses a loop and a series of comparisons to find date range overlaps, where as `my_aorist` uses the `data.table::foverlaps()` function to do this.  The `foverlaps()` function is a C++ implementation of a range join.  There is also a slightly different approach to the two functions calculations of the aoristic weight, but it does not effect execution time or the results (to a very small rounding error).

### Findings

*  `my_aorist()` executes faster when the site/event sample is less the 5,000. At less than 1,000 sites/events, it is about 0.25 seconds faster.
*  `aorist()` executes faster over about 5,000 sites/events. At about 10,000 sites/events it is about 0.25 seconds faster.
*  `my_aorist()` scales poorly with an increase of about 0.55 seconds per 10,000 sites/event
*  `aorist()` scales very well, nearly linearly, with an increase of only about 0.03 seconds per 10,000 sites/events
*  I think it is unlikely for this function to be used for samples of sites/events larger than 5,000, but that extra 0.25 seconds is likely not going to bother anyone.
*  The most interesting finding for my was why the supposedly faster implementation of of `foverlaps()` was scaled so much worse than the supposedly slower implementation of a `for` loop.  The answer is `dplyr`!
*  Without the `dplyr` code, the `my_aorist()` with `foverlaps()` executes much quicker and scales as well as `archSeries::aorist()`
*  Using the `data.table` munge functions and `data.table::foverlaps()` may be the best compromise of computation time.

In the function profiling at the bottom of this analysis, using `profVis`, I think I found the answer to the scaling issue.  The biggest time usage of the `aorist()` is `<=` and `%in%` operator that is executed somewhere within the `data.table` evaluations (not explicit in the function code).  The slowest part of the `my_aorist()` is in the series of `tryCatch` and `doTryCatch` that execute within `mutate_tbl.df()`, presumably within `dplyr::mutate()` that I use. So the `foverlaps()` is not the culprit and leads to the overall quicker execution. This idea was confirmed when I repeated the below code with a version of `my_aorist()` that removed the `dplyr` code; in which the `foverlaps()` code scale equally to `aorist()`, but  0.31 seconds faster on average across a range of 10 to 10,000 sites/events.

### Bottom Line
The `dplyr` data munge in the `my_aorist()` function does not scale well and slows it down pretty dramatically once over 5,000 sites/events. This may be entirely because of my crappy coding, but I'll have try to see if that can improve the time. The `foverlaps()` is lightning fast and without the `dplyr` slowdown makes `my_aorist()` much faster than `aorist()` across a full range of 10 to 10,000 sites/events.  It might be interesting to work `foverlaps()` into the `archSeries::aorist()` function for a speed boost.  Does 0.3 seconds matter to anyone running this analysis; probably not. However, it was fun to find out how these different implementations work.


#### Packages
```{r packages, message=FALSE}
library("ggplot2")
library("profvis")
library("dplyr")
library("tidyr")
library("archSeries")
```


#### `archSeries::aorist()` function

I am testing this as a Global environment function to eliminate any time difference that running it from the installed `archSeries` package may lead to.  (PS - there is no difference, I ran it both ways)

```{r archSeries_aorist, comment=''}
archSeries_aorist <- function (data, weight = 1, start.date = 0, end.date = 2000, 
          bin.width = 100){
  require(data.table)
  data <- data.table(cbind(data, weight)) 
  data <- data[End >= start.date & Start <= end.date]
  data[, `:=`(duration, End - Start)] 
  data[, `:=`(weight.per.unit, weight/duration)]
  breaks <<- seq(start.date, end.date, bin.width)
  labels <- numeric(0) 
  for (i in 1:(length(breaks) - 1)) {
    labels[i] <- paste(breaks[i], breaks[i + 1], sep = "-")
  }
  aorist <- data.table(bin = labels, bin.no = 1:length(labels), 
                       aorist = 0)
  for (i in 1:length(labels)) {
    bin.1 <- breaks[i] 
    bin.2 <- breaks[i + 1] 
    data[, `:=`(assign("a", labels[i]), 0)] 
    data[Start >= bin.1 & Start < bin.2, `:=`(assign("a", 
        labels[i]), (bin.2 - Start) * weight.per.unit)]
    data[End > bin.1 & End <= bin.2, `:=`(assign("a", labels[i]), 
        (End - bin.1) * weight.per.unit)]
    data[Start < bin.1 & End > bin.2, `:=`(assign("a", labels[i]), 
        bin.width * weight.per.unit)]
    data[Start >= bin.1 & End <= bin.2, `:=`(assign("a", 
        labels[i]), as.double(weight))]
    aorist$aorist[i] <- sum(data[, get(labels[i])], na.rm = TRUE)
  }
  aorist
}
```


#### `my_aorist()` function

Here is my implementation of the aorist function using `data.table::foverlaps()` and `dplyr`.

```{r my_aoristic, comment=''}
my_aorist <- function(data, weight = 1, start.date = 0, end.date = 2000, 
                      bin.width = 100, round_int = 4) {
  require(dplyr)
  require(data.table)
  time_steps <- data.frame(
    step  = seq(1:(abs(end.date)/bin.width)), # end not start b/c pos dates
    Start = seq(start.date,(end.date-bin.width), by = bin.width),
    End   = seq((start.date+bin.width),end.date, by = bin.width))
  setDT(time_steps)
  setDT(data)
  setkey(time_steps, Start, End)
  overlap_names <- data.table::foverlaps(data, time_steps, 
                                         type = "any", which = FALSE) %>%
    filter(i.Start != End & i.End != Start)
  aorist <- overlap_names %>%
    data.frame() %>%
    group_by(name) %>%
    mutate(site_count = n(),
           W = (bin.width / (i.End - i.Start)), # SWITCHED b/c of pos dates
           W = round(W,round_int)) %>%
    arrange(name, step) %>%
    group_by(step) %>%
    summarise(step_W = sum(W),
              median_step_W = median(W),
              site_count = n()) %>%
    right_join(., time_steps, by = "step") %>%
    replace_na(list(step_W = 0, site_count = 0, median_step_W = 0)) %>%
    mutate(bin.no = step,
           bin = paste0(Start, "-", End),
           aorist = step_W) %>%
    dplyr::select(bin, bin.no, aorist)
  return(aorist)
}
```



#### Loop over range of sites/events

The `for` loop to execute both functions over a range of sites/events and record the execution time.  I commented out the progress bar code because I don't think `knitr` will appreciate it.

```{r, echo=FALSE}
time_step_size <- 100 # years
seq_begin <- -5000
seq_end   <- 0

max_site_count <- 10000
sim_site_count_seq <- c(10,seq(50,max_site_count,50))

execute_time_results <- matrix(nrow = length(sim_site_count_seq), ncol = 3)
colnames(execute_time_results) <- c("site_count", "aorist", "my_aorist")
# pb <- txtProgressBar(min = 0, max = length(sim_site_count_seq), style = 3, char="*")
for(i in seq_along(sim_site_count_seq)){
  sim_site_count <- sim_site_count_seq[i]
  sites2 <- data.frame(
    name  = seq(1:sim_site_count),
    begin = sample(seq(-5000,-1000, by = 100), sim_site_count, replace = TRUE)) %>%
    mutate(end = begin + sample(seq(100, 3000, by = 100),
                                length(begin), replace = TRUE),
           end = ifelse(end > 0, 0, end)) %>%
    dplyr::select(begin, end, name) %>%
    mutate(begin = abs(begin), end = abs(end)) %>%
    dplyr::rename(Start = end, End = begin)
  
  aor.start.time <- Sys.time()
    aor <- archSeries_aorist(sites2, end.date = 5000, bin.width = 100)
  aor.end.time <- Sys.time()
  my.aor.start.time <- Sys.time()
    my_aor <- my_aorist(sites2, end.date = 5000, bin.width = 100)
  my.aor.end.time <- Sys.time()
  
  aor.execute.time <- aor.end.time - aor.start.time
  my.aor.execute.time <- my.aor.end.time - my.aor.start.time
  execute_time_results[i,1] <- sim_site_count
  execute_time_results[i,2] <- aor.execute.time
  execute_time_results[i,3] <- my.aor.execute.time
  # setTxtProgressBar(pb, i)
}
# close(pb)

```



#### Format and Summarise results

```{r }
execute_time_results <- data.frame(execute_time_results) %>%
  mutate(diff = my_aorist - aorist)

execute_time_summary <- execute_time_results %>%
  summarise(site_count_min = floor(min(site_count)),
            site_count_max = floor(max(site_count)),
            mean_aorist    = round(mean(aorist),3),
            mean_my_aorist = round(mean(my_aorist),3),
            max_diff       = round(min(diff),3), # reverse b/c negative time
            min_diff       = round(max(diff),4), # reverse b/c negative time
            mean_diff      = round(mean(diff),3)) %>%
  t()
print(execute_time_summary)
```


#### Rate of Increase

quick check to see the rates of execution time increases over range of sites/events

```{r}
aor_lm <- lm(aorist ~ site_count, data = execute_time_results)
aor_lm <- (as.numeric(coef(aor_lm))[2]) * 10000
my_aor_lm <- lm(my_aorist ~ site_count, data = execute_time_results)
my_aor_lm <- (as.numeric(coef(my_aor_lm))[2]) * 10000
message(paste0("Initially, my_aoristic is ", round(execute_time_results[2,"diff"],3),
  " seconds faster at ", execute_time_results[2,"site_count"], " sites, is equivalent at 5000 sites, and is ",
  round(execute_time_results[201,"diff"],3), " seconds slower at 10,000 sites. For every 1000 sites added: aoristic = ", round(aor_lm,4), " seconds increase; my_aoristic = ",
               round(my_aor_lm,4), " seconds increase in execution time."))
```



#### Plot results


```{r}
plot_dat <- execute_time_results %>%
  dplyr::select(site_count, aorist, my_aorist) %>%
  gather(model, time, -site_count)

ggplot(plot_dat, aes(x = site_count, y = time, group = model, color = model)) +
  geom_line(size = 1) +
  theme_bw()
```


#### Function profiling

These are best to use the `print()` function to open the profiles in your browser 

```{r function_profiling, comment=''}
profvis({
  aor_profile <- aorist(sites2, end.date = 5000, bin.width = 100)
  # print(aor_profile)
})

profvis({
  my_aor_profile <- my_aorist(sites2, end.date = 5000, bin.width = 100)
  # print(aor_profile)
})

```



### Repeat with no `dplyr`


The same code chunks as above mashed into one chunk with a the only change being the commenting out of `dplyr` code from `my_aorist2()` function.  While this shows the speed increase that is achieved by `foverlaps()`, it is not a fair comparison as the removal of `mutate()` will still need to be replaced by some data munging.  Or, just deal with the half-second of execution time and move along :)

```{r}
my_aorist2 <- function(data, weight = 1, start.date = 0, end.date = 2000, 
                      bin.width = 100, round_int = 4) {
  require(dplyr)
  require(data.table)
  time_steps <- data.frame(
    step  = seq(1:(abs(end.date)/bin.width)), # end not start b/c pos dates
    Start = seq(start.date,(end.date-bin.width), by = bin.width),
    End   = seq((start.date+bin.width),end.date, by = bin.width))
  setDT(time_steps)
  setDT(data)
  setkey(time_steps, Start, End)
  overlap_names <- data.table::foverlaps(data, time_steps, 
                                         type = "any", which = FALSE) %>%
    filter(i.Start != End & i.End != Start)
  aorist <- overlap_names # %>%
#     data.frame() %>%
#     group_by(name) %>%
#     mutate(site_count = n(),
#            W = (bin.width / (i.End - i.Start)), # SWITCHED b/c of pos dates
#            W = round(W,round_int)) %>%
#     arrange(name, step) %>%
#     group_by(step) %>%
#     summarise(step_W = sum(W),
#               median_step_W = median(W),
#               site_count = n()) %>%
#     right_join(., time_steps, by = "step") %>%
#     replace_na(list(step_W = 0, site_count = 0, median_step_W = 0)) %>%
#     mutate(bin.no = step,
#            bin = paste0(Start, "-", End),
#            aorist = step_W) %>%
#     dplyr::select(bin, bin.no, aorist)
  return(aorist)
}

execute_time_results <- matrix(nrow = length(sim_site_count_seq), ncol = 3)
colnames(execute_time_results) <- c("site_count", "aorist", "my_aorist")
# pb <- txtProgressBar(min = 0, max = length(sim_site_count_seq), style = 3, char="*")
for(i in seq_along(sim_site_count_seq)){
  sim_site_count <- sim_site_count_seq[i]
  sites2 <- data.frame(
    name  = seq(1:sim_site_count),
    begin = sample(seq(-5000,-1000, by = 100), sim_site_count, replace = TRUE)) %>%
    mutate(end = begin + sample(seq(100, 3000, by = 100),
                                length(begin), replace = TRUE),
           end = ifelse(end > 0, 0, end)) %>%
    dplyr::select(begin, end, name) %>%
    mutate(begin = abs(begin), end = abs(end)) %>%
    dplyr::rename(Start = end, End = begin)
  
  aor.start.time <- Sys.time()
    aor <- archSeries_aorist(sites2, end.date = 5000, bin.width = 100)
  aor.end.time <- Sys.time()
  my.aor.start.time <- Sys.time()
    my_aor <- my_aorist2(sites2, end.date = 5000, bin.width = 100) # my_aorist2()
  my.aor.end.time <- Sys.time()
  
  aor.execute.time <- aor.end.time - aor.start.time
  my.aor.execute.time <- my.aor.end.time - my.aor.start.time
  execute_time_results[i,1] <- sim_site_count
  execute_time_results[i,2] <- aor.execute.time
  execute_time_results[i,3] <- my.aor.execute.time
  # setTxtProgressBar(pb, i)
}
# close(pb)

execute_time_results <- data.frame(execute_time_results) %>%
  mutate(diff = my_aorist - aorist)

execute_time_summary <- execute_time_results %>%
  summarise(site_count_min = floor(min(site_count)),
            site_count_max = floor(max(site_count)),
            mean_aorist    = round(mean(aorist),3),
            mean_my_aorist = round(mean(my_aorist),3),
            max_diff       = round(min(diff),3), # reverse b/c negative time
            min_diff       = round(max(diff),4), # reverse b/c negative time
            mean_diff      = round(mean(diff),3)) %>%
  t()
print(execute_time_summary)

aor_lm <- lm(aorist ~ site_count, data = execute_time_results)
aor_lm <- (as.numeric(coef(aor_lm))[2]) * 10000
my_aor_lm <- lm(my_aorist ~ site_count, data = execute_time_results)
my_aor_lm <- (as.numeric(coef(my_aor_lm))[2]) * 10000
message(paste0("Initially, my_aoristic is ", round(execute_time_results[2,"diff"],3),
  " seconds faster at ", execute_time_results[2,"site_count"], " sites, is equivalent at 5000 sites, and is ",
  round(execute_time_results[201,"diff"],3), " seconds slower at 10,000 sites. For every 1000 sites added: aoristic = ", round(aor_lm,4), " seconds increase; my_aoristic = ",
               round(my_aor_lm,4), " seconds increase in execution time."))

plot_dat <- execute_time_results %>%
  dplyr::select(site_count, aorist, my_aorist) %>%
  gather(model, time, -site_count)

ggplot(plot_dat, aes(x = site_count, y = time, group = model, color = model)) +
  geom_line(size = 1) +
  theme_bw()

```













